---
title: "Ebola Cases over the years 2014-2016 (Assessment_AW)"
author: "Anna Wyss"
format: html
editor: visual
date: today
fromat: 
  html:
    embeded-resources: true
---

```{r}
#| output: false
#| echo: false
#| message: false
#| warning: false
# load libraries
library(here)
library(tidyverse)
library(lubridate)
library(unibeCols)
library(htmlTable)
library(knitr)
```

```{r}
#| echo: false
#| message: false
#| warning: false
ebola <- read_csv("/Users/anna/Desktop/Uni:Learing/Visualization_R/ProjectsInR/data/raw/ebola.csv")

#create subset for each year (2014, 2015, 2016) and over the whole timeperiod for the countries Liberia, Sierra Leone, United Kingdom, Spain
ebola_date_2014 <- ebola %>% filter(Date >= ymd("2014-08-29") & Date <= ymd("2014-12-31")
                                       & (Country == "Liberia" | Country == "Sierra Leone" | Country == "United Kingdom" | Country == "Spain"))


ebola_date_2015 <- ebola %>% filter(Date >= ymd("2015-01-01") & Date <= ymd("2015-12-31")
                                       & (Country == "Liberia" | Country == "Sierra Leone" | Country == "United Kingdom" | Country == "Spain"))


ebola_date_2016 <- ebola %>% filter(Date >= ymd("2016-01-01") & Date <= ymd("2016-12-31")
                                       & (Country == "Liberia" | Country == "Sierra Leone" | Country == "United Kingdom" | Country == "Spain"))

ebola_date_total <- ebola %>% filter((Country == "Liberia" | Country == "Sierra Leone" | Country == "United Kingdom" | Country == "Spain"))
```

## Aim of the Report
In this report ebola labratory-confirmed cases in Liberia, Sierra Leone, the United Kingdom and Spain will be investigated for the years 2014 - 2016. The ebola data set is downloaded form the [Projects in R Repository](https://github.com/ISPMBern/projects-in-R/tree/main/data). 

## Labratory-confirmed Cases over the Years
### Year 2014
Please note that the data collection only started in August 2014. In @fig-ebola14 the development from end of August 2014 to December 2014 is displayed. A rapid increase in cases in Sierra Leone and a slight increase in Liberia and no increase in Spain is observable. For the United Kingdome no ebola infection was reported during this time period. 
```{r}
#| label: fig-ebola14
#| fig-cap: "Confirmed Ebola Cases (Liberia, Sierra Leone, United Kingdom, Spain) in 2014. Please note that there no ebola case in the United Kingdome in 2014."
#| fig-width: 8
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false

country_shapes <- c("Liberia" = 0, 
                    "Sierra Leone" = 1, 
                    "United Kingdom" = 2, 
                    "Spain" = 8)

ggplot(data = ebola_date_2014, 
       mapping = aes(x = Date, y = Cum_conf_cases, color = Country, shape = Country)) + 
  geom_point(size = 1, stroke = 1) +
  scale_color_brewer(palette = "Accent") +
  scale_shape_manual(values = country_shapes) +
  labs(title = "Ebola Cases (Liberia, Sierra Leone, Spain) in 2014",
       x = "Date", 
       y = "Confirmed Cases (cumulative)")

```
```{r}
#| label: tbl-ebola14
#| tbl-cap: "Exact numbers of labratory-confirmed cases in each country on 31.12.2014."
#| fig-width: 8
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
tbl14 <- ebola_date_2014 %>%
  group_by(Country) %>%
  summarize(Cum_conf_cases = max(Cum_conf_cases))

kable(tbl14, caption = "", col.names = c("Country", "Total 31.12.2014"))
```


### Year 2015
In @fig-ebola15 a stabilization at a high number of laboratory-confirmed cases in Sierra Leone and stabilization at a mediate number of cases in Liberia can be observed. Again, there are no cases in Spain and also no in the United Kingdom.
```{r}
#| label: fig-ebola15
#| fig-cap: "Confirmed Ebola Cases (Liberia, Sierra Leone, United Kingdom, Spain) in 2015."
#| fig-width: 8
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
ggplot(data = ebola_date_2015, 
       mapping = aes(x = Date, y = Cum_conf_cases, color = Country, shape = Country)) + 
  geom_point(size = 1, stroke = 1) +
  scale_color_brewer(palette = "Accent") +
  scale_shape_manual(values = country_shapes) +
  labs(title = "Ebola Cases (Liberia, Sierra Leone, United Kingdom, Spain) in 2015",
       x = "Date", 
       y = "Confirmed Cases (cumulative)")

```

```{r}
#| label: tbl-ebola15
#| tbl-cap: "Exact numbers of labratory-confirmed cases in each country on 31.12.2015."
#| fig-width: 8
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
tbl15 <- ebola_date_2015 %>%
  group_by(Country) %>%
  summarize(Cum_conf_cases = max(Cum_conf_cases))

kable(tbl15, caption = "", col.names = c("Country", "Total 31.12.2015"))
```


### Year 2016
Only very little data points are accessable for 2016. Please consider @fig-ebola16.
```{r}
#| label: fig-ebola16
#| fig-cap: "Confirmed Ebola Cases (Liberia, Sierra Leone, United Kingdom, Spain) in 2016."
#| fig-width: 8
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
ggplot(data = ebola_date_2016, 
       mapping = aes(x = Date, y = Cum_conf_cases, color = Country, shape = Country)) + 
  geom_point(size = 1, stroke = 1) +
  scale_color_brewer(palette = "Accent") +
  scale_shape_manual(values = country_shapes) +
  labs(title = "Ebola Cases (Liberia, Sierra Leone, United Kingdom, Spain) in 2016",
       x = "Date", 
       y = "Confirmed Cases (cumulative)")


```

```{r}
#| label: tbl-ebola16
#| tbl-cap: "Exact numbers of labratory-confirmed cases in each country on 23.03.2016."
#| fig-width: 8
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
tbl16 <- ebola_date_2016 %>%
  group_by(Country) %>%
  summarize(Cum_conf_cases = max(Cum_conf_cases))

kable(tbl16, caption = "", col.names = c("Country", "Total 23.03.2016"))
```
### Timeperiod 2014 - 2016
To get a more comprehensive picture of labratory-confirmed cases, please consider @fig-ebolatotal and @fig-colebolatotal. There, the labratory-confirmed cases over the timeperiod August 2014 - March 2016 are displayed.
```{r}
#| label: fig-ebolatotal
#| fig-cap: "Confirmed Ebola Cases (Liberia, Sierra Leone, United Kingdom, Spain) from 2014-2016."
#| fig-width: 8
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
ggplot(data = ebola_date_total, 
       mapping = aes(x = Date, y = Cum_conf_cases, color = Country, shape = Country)) + 
  geom_point(size = 1, stroke = 1) +
  scale_color_brewer(palette = "Accent") +
  scale_shape_manual(values = country_shapes) +
  labs(title = "Ebola Cases (Liberia, Sierra Leone, United Kingdom, Spain) from 2014-2016",
       x = "Date", 
       y = "Confirmed Cases (cumulative)")
```

```{r}
#| label: fig-colebolatotal
#| fig-cap: "Ebola Cases (Liberia, Sierra Leone, United Kingdom, Spain) from 2014-2016."
#| fig-width: 8
#| fig-height: 4
#| echo: false
#| message: false
#| warning: false
ggplot(data = ebola_date_total, 
       mapping = aes(x = Date, y = Cum_conf_cases, color = Country, shape = Country)) + 
  geom_col(size = 0.3) +
  scale_color_brewer(palette = "Accent") +
  labs(title = "Ebola Cases (Liberia, Sierra Leone, United Kingdom, Spain) from 2014-2016",
       x = "Date", 
       y = "Confirmed Cases (cumulative)")
```
## Conclusion
In this report it can be obseverd that after an ebola outbreak (>1), there is an rapide increase in labratory-confirmed cases which stabilizes over the following years. It should be the main aim to prevent an ebola outbrake at any cost. 

